Possible Inconsistency in Maximum Likelihood Calculation of Ancestral States

### 2017-04-28 joint_inf.tex
fixing (partial) log likelihoods

We are interested in the properties of the objective function being maximized in \eqref{eq:site_pattern_likelihood} as data becomes plentiful.
%EM As described IRL, a little strange to be redefining (9) with an extra variable. After reading the below, I think it makes most sense to just refer to (9) indirectly but make this a stand-alone definition.
Let
$$
L_\nCols''(\tau, t; \fullAlignment, \fullAncestralSplitPartitions(\tau, t)) = \prod_{j=1}^{\nSiteSplits} \ \left[P(\siteSplitRV=\siteSplit_j \mid \tau, t) \cdot P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)}
$$
be the final product in \eqref{eq:site_pattern_likelihood}.
Assume our $\nCols$ observations were generated from a model with parameters $(\tau^*, t^*)$.
We have
\begin{align}
    \frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment, \fullAncestralSplitPartitions(\tau,t))
        &= \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot  \log P(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_j \mid \tau, t) \\
        &= \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot [\log P(\siteSplitRV=\siteSplit_j \mid \tau, t) +
            \log P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)]
\end{align}
so that, in the limit of infinite data,
\begin{equation}
%EM did you mean to drop the eta from the first term of this equation?
    \frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment, \fullAncestralSplitPartitions(\tau, t)) \rightarrow \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot [\log P(\siteSplitRV=\siteSplit_j \mid \tau, t) + \log P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)]. \label{eq:site_pattern_profile_likelihood_mean}
\end{equation}
Define the the divergence quantity
$$
\shannonDivergence_{\tau^*,t^*}(\tau,t) = \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log P(\siteSplitRV=\siteSplit_j \mid \tau, t),
$$
and the partial log likelihood
$$
%EM IMHO \ell^p is so standard that it's confusing to see it in another setting. Perhaps \tilde{\ell}?
\tilde{\ell}_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t) = \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log P(\ancestralSplitRV=\xi_j \mid \siteSplit_j, \tau, t)
$$
so that \eqref{eq:site_pattern_profile_likelihood_mean} becomes
\begin{equation}
    \label{eq:log_likelihood_simplified}
    \ell_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t) = \shannonDivergence_{\tau^*,t^*}(\tau,t) + \tilde{\ell}_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t).
\end{equation}
Recall $(\tau, t)$ are estimands and $(\tau^*, t^*)$ are fixed, generating parameters.
For data generated from $(\tau^*, t^*)$, if
\begin{equation}
\label{eq:inconsistency_inequality}
%EM Suggest \hat{\fullAncestralSplitPartitions}_1, etc.
\ell_{\tau^*,t^*}(\hat{\fullAncestralSplitPartitions}_1,\tau',\hat{t}_1) > \ell_{\tau^*,t^*}(\hat{\fullAncestralSplitPartitions}_2,\tau^*,\hat{t}_2)
\end{equation}
for $\tau'\neq\tau^*$ with $\{\hat{t}_1,\hat{\fullAncestralSplitPartitions}_1\}$ and $\{\hat{t}_2,\hat{\fullAncestralSplitPartitions}_2\}$ estimated by maximizing \eqref{eq:profile_likelihood}, we have an inconsistency.
By Gibbs' inequality, $\shannonDivergence_{\tau^*,t^*}(\tau,t)$ is maximized when $(\tau,t) = (\tau^*,t^*)$, so any potential example of inconsistency must overcome that difference with substantially greater partial likelihoods on the non-generating topology.
In the next section, we show that such examples exist.

@@

Our approach will be to seek bounds for the terms inside the maximization in \eqref{eq:log_likelihood_simplified}.
We obtain an upper bound for the joint maximum of \eqref{eq:log_likelihood_simplified} as
$$
\max_{t,\fullAncestralSplitPartitions(\tau,t)} \ \ell_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t) \le
    \shannonDivergence_{\tau^*,t^*}(\tau^*,t^*)
+ \tilde{\ell}_{\tau^*,t^*}(\hat{\fullAncestralSplitPartitions},\tau,\hat{t})
$$
using Gibbs's inequality
$$
\shannonDivergence_{\tau^*,t^*}(\tau,t) \le \shannonDivergence_{\tau^*,t^*}(\tau^*,t^*)
$$
and
$$
\{\hat{t},\hat{\fullAncestralSplitPartitions}\} = \argmax_{t,\fullAncestralSplitPartitions(\tau,t)} \ \tilde{\ell}_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t).
$$
Similarly, we have a lower bound as
$$
\max_{t,\fullAncestralSplitPartitions(\tau,t)} \ \ell_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t) \ge
    \shannonDivergence_{\tau^*,t^*}(\tau,\hat{t})
+ \tilde{\ell}_{\tau^*,t^*}(\hat{\fullAncestralSplitPartitions},\tau,\hat{t}).
$$

@@

%das: did this on paper, but can write it out.
\textbf{Justification for estimand:} Consider the Farris topology with arbitrary branch lengths, i.e., $\{x_1,y_1,x_2,y_2,w\}$ counting from the top left.
The likelihood $\ell(x_1,y_1,x_2,y_2,w)$ can be written as a function of $\{x_1x_2, x_1+x_2, y_1y_2, y_1+y_2, w\}$ only.
In this case, $x_1$ and $x_2$ can be exchanged for one another without changing the value of $\ell$, and any maximum will require $x_1=x_2$.
A similar argument can be made for $y_1$ and $y_2$.
For the Felsenstein topolgy this is not the case, but, since we are interested in a lower bound for this topology, we will simplify the objective function by constraining $x_1=x_2$ and $y_1=y_2$ similarly.

@@

    \ell_{\tau_1,\{x,y,y\}}(\fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}),\tau_1,\{x',y',w'\})

@@

The other two possibilties of ancestral state split for the likelihood of the site split $\{1,3\}$ admit similar simplifications.
Call
$$
a_{1}' = a_{1}-2p_{13},
$$
$$
a_{2}' = a_{2}+2p_{13},
$$
$$
a_{3}' = a_{3}-2p_{13},
$$
$$
a_{4}' = a_{4}+2p_{13}.
$$
Then, for these cases,
\begin{align*}
    \ell_{\tau_1,\{x,y,y\}}(\fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}),\tau_1,\{x',y',w'\})
    &\le      p_{0}  \cdot\log(p_{0}) \\
    &\qquad + p_{13} \cdot\log(p_{13}) \\
    &\qquad + a_{1}'\cdot\log\frac{2a_{1}'}{a_{1}'+a_{2}'} \\
    &\qquad + a_{2}'\cdot\log\frac{2a_{2}'}{a_{1}'+a_{2}'} \\
    &\qquad + a_{3}\cdot\log\frac{2a_{3}}{a_{3}+a_{4}} \\
    &\qquad + a_{4}\cdot\log\frac{2a_{4}}{a_{3}+a_{4}} \\
    &\qquad + \log(2) \\
    &\qquad := L^{2}_{\tau_1,\tau_1}(\mathbf{p}_{xy});
\end{align*}
\begin{align*}
    \ell_{\tau_1,\{x,y,y\}}(\fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}),\tau_1,\{x',y',w'\})
    &\le      p_{0}  \cdot\log(p_{0}) \\
    &\qquad + p_{13} \cdot\log(p_{13}) \\
    &\qquad + a_{1}\cdot\log\frac{2a_{1}}{a_{1}+a_{2}} \\
    &\qquad + a_{2}\cdot\log\frac{2a_{2}}{a_{1}+a_{2}} \\
    &\qquad + a_{3}'\cdot\log\frac{2a_{3}'}{a_{3}'+a_{4}'} \\
    &\qquad + a_{4}'\cdot\log\frac{2a_{4}'}{a_{3}'+a_{4}'} \\
    &\qquad + \log(2) \\
    &\qquad := L^{3}_{\tau_1,\tau_1}(\mathbf{p}_{xy}).
\end{align*}
All bounds are functions only of $x,y$ through the true generating probabilities $\mathbf{p}_{xy}$.

@@

We construct a similar lower bound on the Felsenstein partial likelihood.
Replace all $(1+w)$ terms with $(1-w)$ and focus just on the partial likelihood.
Then if $(1+x)(1-y) > (1-x)(1+y)$,
\begin{align*}
    \tilde{\ell}_{\tau_1,\{x,y,y\}}(\fullAncestralSplitPartitions(\tau_2,\{x',y',w'\}),\tau_2,\{x',y',w'\})
    &=      p_{0}  \cdot\log((1+x')^2   (1+w')(1+y')^2) \\
    &\qquad + p_{1}  \cdot\log((1+x')^2   (1-w')(1+y')(1-y')) \\
    &\qquad + p_{2}  \cdot\log((1+x')^2   (1-w')(1+y')(1-y')) \\
    &\qquad + p_{3}  \cdot\log((1+x')^2   (1-w')(1+y')(1-y')) \\
    &\qquad + p_{12} \cdot\log((1+x')^2   (1-w')(1+y')^2) \\
    &\qquad + p_{123}\cdot\log((1+x')^2   (1-w')(1+y')(1-y'))\\
    &\qquad + p_{13} \cdot\log((1+x')^2   (1-w')(1-y')^2) \\
    &\qquad + p_{23} \cdot\log((1+x')^2   (1-w')(1-y')^2),
\end{align*}
otherwise
\begin{align*}
    \tilde{\ell}_{\tau_1,\{x,y,y\}}(\fullAncestralSplitPartitions(\tau_2,\{x',y',w'\}),\tau_2,\{x',y',w'\})
    &=      p_{0}  \cdot\log((1+x')^2    (1+w')(1+y')^2) \\
    &\qquad + p_{1}  \cdot\log((1+x')(1-x')(1-w')(1+y')^2) \\
    &\qquad + p_{2}  \cdot\log((1+x')(1-x')(1-w')(1+y')^2) \\
    &\qquad + p_{3}  \cdot\log((1+x')(1-x')(1-w')(1+y')^2) \\
    &\qquad + p_{12} \cdot\log((1+x')^2    (1-w')(1+y')^2) \\
    &\qquad + p_{123}\cdot\log((1+x')(1-x')(1-w')(1+y')^2)\\
    &\qquad + p_{13} \cdot\log((1-x')^2    (1-w')(1+y')^2) \\
    &\qquad + p_{23} \cdot\log((1-x')^2    (1-w')(1+y')^2).
\end{align*}
Letting
$$
b = p_{1}+p_{2}+p_{3}+p_{123}
$$
yields
\begin{align*}
    \tilde{\ell}_{\tau_1,\{x,y,y\}}(\fullAncestralSplitPartitions(\tau_2,\{x',y',w'\}),\tau_2,\{x',y',w'\})
    &=      2\cdot\log(1+x') \\
    &\qquad + (2-b)  \cdot\log(1+y') \\
    &\qquad + b      \cdot\log(1-y') \\
    &\qquad + p_{0}\cdot\log(1+w') \\
    &\qquad + (1-p_{0})\cdot\log(1-w')
\end{align*}
for the first lower bound, but, upon maximizing, we obtain the same value for both cases, namely
\begin{align*}
    \tilde{\ell}_{\tau_1,\{x,y,y\}}(\fullAncestralSplitPartitions(\tau_2,\{x',y',w'\}),\tau_2,\{x',y',w'\})
    &=      2\cdot\log(2) \\
    &\qquad + (2-b)  \cdot\log(2-b) \\
    &\qquad + b      \cdot\log(b) \\
    &\qquad + p_{0}\cdot\log(2p_{0}) \\
    &\qquad + (1-p_{0})\cdot\log(2(1-p_{0}))
\end{align*}
maximized at $\{1,1-b,1-2p_{0}\}$.
Our lower bound is
\begin{align*}
    L_{\tau_1,\tau_2}(\mathbf{p}_{xy}) &:= \shannonDivergence_{\tau_1,\{x,y,y\}}(\tau_2,\{1,1-b,1-2p_{0}\}) \\
                                           &+ 2\cdot\log(2) + (2-b)  \cdot\log(2-b) + b      \cdot\log(b) + p_{0}\cdot\log(2p_{0}) + (1-p_{0})\cdot\log(2(1-p_{0})).
\end{align*}

>>>>

We are interested in the properties of the objective function being maximized in \eqref{eq:site_pattern_likelihood} as data becomes plentiful.
%EM As described IRL, a little strange to be redefining (9) with an extra variable. After reading the below, I think it makes most sense to just refer to (9) indirectly but make this a stand-alone definition.
Let
$$
L_\nCols''(\tau, t; \fullAlignment, \fullAncestralSplitPartitions(\tau, t)) = \prod_{j=1}^{\nSiteSplits} \ \left[P(\siteSplitRV=\siteSplit_j \mid \tau, t) \cdot P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)}
$$
be the final product in \eqref{eq:site_pattern_likelihood}.
Assume our $\nCols$ observations were generated from a model with parameters $(\tau^*, t^*)$.
We have
\begin{align}
    \frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment, \fullAncestralSplitPartitions(\tau,t))
        &= \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot  \log P(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_j \mid \tau, t) \\
        &= \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot [\log P(\siteSplitRV=\siteSplit_j \mid \tau, t) +
            \log P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)]
\end{align}
so that, in the limit of infinite data,
\begin{equation}
%EM did you mean to drop the eta from the first term of this equation?
    \frac{1}{\nCols} \log L_\nCols''(\tau, t; \fullAlignment, \fullAncestralSplitPartitions(\tau, t)) \rightarrow \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot [\log P(\siteSplitRV=\siteSplit_j \mid \tau, t) + \log P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)]. \label{eq:site_pattern_profile_likelihood_mean}
\end{equation}
Define the the divergence quantity
$$
\shannonDivergence_{\tau^*,t^*}(\tau,t) = \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log P(\siteSplitRV=\siteSplit_j \mid \tau, t),
$$
and the partial log likelihood
$$
%EM IMHO \ell^p is so standard that it's confusing to see it in another setting. Perhaps \tilde{\ell}?
\tilde{\ell}_{\tau^*,t^*}(\tau, t; \fullAncestralSplitPartitions(\tau,t)) = \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log P(\ancestralSplitRV=\xi_j \mid \siteSplit_j, \tau, t)
$$
so that \eqref{eq:site_pattern_profile_likelihood_mean} becomes
\begin{equation}
    \label{eq:log_likelihood_simplified}
    \ell_{\tau^*,t^*}(\tau, t; \fullAncestralSplitPartitions(\tau,t)) = \shannonDivergence_{\tau^*,t^*}(\tau,t) + \tilde{\ell}_{\tau^*,t^*}(\tau, t; \fullAncestralSplitPartitions(\tau,t)).
\end{equation}
Recall $(\tau, t)$ are estimands and $(\tau^*, t^*)$ are fixed, generating parameters.
For data generated from $(\tau^*, t^*)$, if
\begin{equation}
\label{eq:inconsistency_inequality}
%EM Suggest \hat{\fullAncestralSplitPartitions}_1, etc.
\ell_{\tau^*,t^*}(\tau', \hat{t}_1; \hat{\fullAncestralSplitPartitions}_1) > \ell_{\tau^*,t^*}(\tau^*, \hat{t}_2; \hat{\fullAncestralSplitPartitions}_2)
\end{equation}
for $\tau'\neq\tau^*$ with $\{\hat{t}_1,\hat{\fullAncestralSplitPartitions}_1\}$ and $\{\hat{t}_2,\hat{\fullAncestralSplitPartitions}_2\}$ estimated by maximizing \eqref{eq:profile_likelihood}, we have an inconsistency.
By Gibbs' inequality, $\shannonDivergence_{\tau^*,t^*}(\tau,t)$ is maximized when $(\tau,t) = (\tau^*,t^*)$, so any potential example of inconsistency must overcome that difference with substantially greater partial likelihoods on the non-generating topology.
In the next section, we show that such examples exist.

@@

Our approach will be to seek bounds for the terms inside the maximization in \eqref{eq:log_likelihood_simplified}.
We obtain an upper bound for the joint maximum of \eqref{eq:log_likelihood_simplified} as
$$
\max_{t,\fullAncestralSplitPartitions(\tau,t)} \ \ell_{\tau^*,t^*}(\tau, t; \fullAncestralSplitPartitions(\tau,t)) \le
    \shannonDivergence_{\tau^*,t^*}(\tau^*,t^*)
    + \tilde{\ell}_{\tau^*,t^*}(\tau, \hat{t}; \hat{\fullAncestralSplitPartitions})
$$
using Gibbs's inequality
$$
\shannonDivergence_{\tau^*,t^*}(\tau,t) \le \shannonDivergence_{\tau^*,t^*}(\tau^*,t^*)
$$
and
$$
\{\hat{t},\hat{\fullAncestralSplitPartitions}\} = \argmax_{t,\fullAncestralSplitPartitions(\tau,t)} \ \tilde{\ell}_{\tau^*,t^*}(\tau, t; \fullAncestralSplitPartitions(\tau,t)).
$$
Similarly, we have a lower bound as
$$
\max_{t,\fullAncestralSplitPartitions(\tau,t)} \ \ell_{\tau^*,t^*}(\tau, t; \fullAncestralSplitPartitions(\tau,t)) \ge
    \shannonDivergence_{\tau^*,t^*}(\tau,\hat{t})
    + \tilde{\ell}_{\tau^*,t^*}(\tau, \hat{t}; \hat{\fullAncestralSplitPartitions}).
$$

@@

%das: did this on paper, but can write it out.
\textbf{Justification for estimand:} Consider the Farris topology with arbitrary branch lengths, i.e., $\tilde{t}=\{x_1,y_1,x_2,y_2,w\}$ counting from the top left.
The log likelihood $\ell_{\tau_1,t^*}(\tau_1, \tilde{t}; \fullAncestralSplitPartitions(\tau_1,\tilde{t}))$ can be written as a function of $\{x_1x_2, x_1+x_2, y_1y_2, y_1+y_2, w\}$ only.
In this case, $x_1$ and $x_2$ can be exchanged for one another without changing the value of $\ell$, and any maximum will require $x_1=x_2$.
A similar argument can be made for $y_1$ and $y_2$.
For the Felsenstein topolgy this is not the case, but, since we are interested in a lower bound for this topology, we will simplify the objective function by constraining $x_1=x_2$ and $y_1=y_2$ similarly.

@@

    \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}))

@@

The other two possibilties of ancestral state split for the likelihood of the site split $\{1,3\}$ admit similar simplifications.
Call
$$
a_{1}' = a_{1}-2p_{13},
$$
$$
a_{2}' = a_{2}+2p_{13},
$$
$$
a_{3}' = a_{3}-2p_{13},
$$
$$
a_{4}' = a_{4}+2p_{13}.
$$
Then, for these cases,
\begin{align*}
    \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}))
    &\le      p_{0}  \cdot\log(p_{0}) \\
    &\qquad + p_{13} \cdot\log(p_{13}) \\
    &\qquad + a_{1}'\cdot\log\frac{2a_{1}'}{a_{1}'+a_{2}'} \\
    &\qquad + a_{2}'\cdot\log\frac{2a_{2}'}{a_{1}'+a_{2}'} \\
    &\qquad + a_{3}\cdot\log\frac{2a_{3}}{a_{3}+a_{4}} \\
    &\qquad + a_{4}\cdot\log\frac{2a_{4}}{a_{3}+a_{4}} \\
    &\qquad + \log(2) \\
    &\qquad := L^{2}_{\tau_1,\tau_1}(\mathbf{p}_{xy});
\end{align*}
\begin{align*}
    \ell_{\tau_1,\{x,y,y\}}(\tau_1, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_1,\{x',y',w'\}))
    &\le      p_{0}  \cdot\log(p_{0}) \\
    &\qquad + p_{13} \cdot\log(p_{13}) \\
    &\qquad + a_{1}\cdot\log\frac{2a_{1}}{a_{1}+a_{2}} \\
    &\qquad + a_{2}\cdot\log\frac{2a_{2}}{a_{1}+a_{2}} \\
    &\qquad + a_{3}'\cdot\log\frac{2a_{3}'}{a_{3}'+a_{4}'} \\
    &\qquad + a_{4}'\cdot\log\frac{2a_{4}'}{a_{3}'+a_{4}'} \\
    &\qquad + \log(2) \\
    &\qquad := L^{3}_{\tau_1,\tau_1}(\mathbf{p}_{xy}).
\end{align*}
All bounds are functions only of $x,y$ through the true generating probabilities $\mathbf{p}_{xy}$.

@@

We construct a similar lower bound on the Felsenstein partial likelihood.
Replace all $(1+w)$ terms with $(1-w)$ and focus just on the partial likelihood.
Then if $(1+x)(1-y) > (1-x)(1+y)$,
\begin{align*}
    \tilde{\ell}_{\tau_1,\{x,y,y\}}(\tau_2, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_2,\{x',y',w'\}))
    &=      p_{0}  \cdot\log((1+x')^2   (1+w')(1+y')^2) \\
    &\qquad + p_{1}  \cdot\log((1+x')^2   (1-w')(1+y')(1-y')) \\
    &\qquad + p_{2}  \cdot\log((1+x')^2   (1-w')(1+y')(1-y')) \\
    &\qquad + p_{3}  \cdot\log((1+x')^2   (1-w')(1+y')(1-y')) \\
    &\qquad + p_{12} \cdot\log((1+x')^2   (1-w')(1+y')^2) \\
    &\qquad + p_{123}\cdot\log((1+x')^2   (1-w')(1+y')(1-y'))\\
    &\qquad + p_{13} \cdot\log((1+x')^2   (1-w')(1-y')^2) \\
    &\qquad + p_{23} \cdot\log((1+x')^2   (1-w')(1-y')^2),
\end{align*}
otherwise
\begin{align*}
    \tilde{\ell}_{\tau_1,\{x,y,y\}}(\tau_2, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_2,\{x',y',w'\}))
    &=      p_{0}  \cdot\log((1+x')^2    (1+w')(1+y')^2) \\
    &\qquad + p_{1}  \cdot\log((1+x')(1-x')(1-w')(1+y')^2) \\
    &\qquad + p_{2}  \cdot\log((1+x')(1-x')(1-w')(1+y')^2) \\
    &\qquad + p_{3}  \cdot\log((1+x')(1-x')(1-w')(1+y')^2) \\
    &\qquad + p_{12} \cdot\log((1+x')^2    (1-w')(1+y')^2) \\
    &\qquad + p_{123}\cdot\log((1+x')(1-x')(1-w')(1+y')^2)\\
    &\qquad + p_{13} \cdot\log((1-x')^2    (1-w')(1+y')^2) \\
    &\qquad + p_{23} \cdot\log((1-x')^2    (1-w')(1+y')^2).
\end{align*}
Letting
$$
b = p_{1}+p_{2}+p_{3}+p_{123}
$$
yields
\begin{align*}
    \tilde{\ell}_{\tau_1,\{x,y,y\}}(\tau_2, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_2,\{x',y',w'\}))
    &=      2\cdot\log(1+x') \\
    &\qquad + (2-b)  \cdot\log(1+y') \\
    &\qquad + b      \cdot\log(1-y') \\
    &\qquad + p_{0}\cdot\log(1+w') \\
    &\qquad + (1-p_{0})\cdot\log(1-w')
\end{align*}
for the first lower bound, but, upon maximizing, we obtain the same value for both cases, namely
\begin{align*}
    \tilde{\ell}_{\tau_1,\{x,y,y\}}(\tau_2, \{x',y',w'\}; \fullAncestralSplitPartitions(\tau_2,\{x',y',w'\}))
    &=      2\cdot\log(2) \\
    &\qquad + (2-b)  \cdot\log(2-b) \\
    &\qquad + b      \cdot\log(b) \\
    &\qquad + p_{0}\cdot\log(2p_{0}) \\
    &\qquad + (1-p_{0})\cdot\log(2(1-p_{0}))
\end{align*}
maximized at $\{1,1-b,1-2p_{0}\}$.
Our lower bound is
\begin{align*}
    L_{\tau_1,\tau_2}(\mathbf{p}_{xy}) &:= \shannonDivergence_{\tau_1,\{x,y,y\}}(\tau_2,\{1,1-b,1-2p_{0}\}) \\
                                           &+ 2\cdot\log(2) + (2-b)  \cdot\log(2-b) + b      \cdot\log(b) + p_{0}\cdot\log(2p_{0}) + (1-p_{0})\cdot\log(2(1-p_{0})).
\end{align*}
